---
output: md_document
---

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/r2rtf)](https://CRAN.R-project.org/package=r2rtf)
[![Codecov test coverage](https://codecov.io/gh/Merck/r2rtf/branch/master/graph/badge.svg)](https://codecov.io/gh/Merck/r2rtf?branch=master)
[![R build status](https://github.com/Merck/r2rtf/workflows/R-CMD-check/badge.svg)](https://github.com/Merck/r2rtf/actions)
[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/r2rtf)](https://cran.r-project.org/package=r2rtf)
<!-- badges: end -->
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}
pkgname <- "r2rtf"
```

# Overview 

Create RTF table and figure with flexible format.

## Installation

You can install the package via CRAN:

```r
install.packages("r2rtf")
```

Or, install from GitHub for development version:

```r
remotes::install_github("Merck/r2rtf")
```

## Highlighted Features 

The R package `r pkgname` provided flexibility to provide features below:

* Necessary options to create highly customized RTF table and figure. 
* Simple to use parameters and data structure.
    + Customized column header: split by "|". 
    + Three required parameters for the output tables (data, filename, column relative width).
    + Flexible and detail control of table structure. 
* Format control in cell, row, column and table level for: 
    + Border Type: single, double, dash, dot, etc.
    + Alignment: left, right, center, decimal.
    + Column width. 
    + Text appearance: **bold**, *italics*, ~~strikethrough~~, underline and any combinations.
    + Font size. 
    + Text and border color (`r length(colors())` different colors named in `color()` function).
    + Special characters: any character in UTF-8 encoding (e.g. Greek, Symbol, Chinese, Japanese, Korean).
* Append several tables into one file. 
* Pagination.
* Built in raw data for validation.

## Quick Example 

```{r, eval = FALSE}
library(dplyr)
library(r2rtf)
head(iris) %>% rtf_body() %>%                       # Step 1 Add attributes 
               rtf_encode() %>%                     # Step 2 Convert attributes to RTF encode 
               write_rtf(file = "ex-tbl.rtf")       # Step 3 Write to a .rtf file 
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("vignettes/pdf/ex-tbl.pdf")
```

## Example Efficacy Table

- [Source code](https://merck.github.io/r2rtf/articles/example-efficacy.html)

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("vignettes/pdf/efficacy_example.pdf")
```

## Example Safety Table

- [Source code](https://merck.github.io/r2rtf/articles/example-ae-summary.html)

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("vignettes/pdf/ae_example.pdf")
```
