---
output: md_document
---

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/r2rtf)](https://CRAN.R-project.org/package=r2rtf)
[![Codecov test coverage](https://codecov.io/gh/Merck/r2rtf/branch/master/graph/badge.svg)](https://codecov.io/gh/Merck/r2rtf?branch=master)
[![R build status](https://github.com/Merck/r2rtf/workflows/R-CMD-check/badge.svg)](https://github.com/Merck/r2rtf/actions)
<!-- badges: end -->
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}
pkgname = "r2rtf"
```

# Overview 

Create RTF table and figure with flexible format.

## Installation

1. Download the package in this code repository.
2. Unzip the file.
3. Open the *.Rproj file by Rstudio.
4. In Rstudio console, run `devtools::install()`.

## Highlighted Features 

The R package `r pkgname` provided flexibility to provide features below:

* Necessary options to create highly customized RTF table and figure. 
* Simple to use parameters and data structure.
    + Customized column header: split by "|". 
    + Three required parameters for the output tables (data, filename, column relative width).
    + Flexible and detail control of table structure. 
* Format control in cell, row, column and table level for: 
    + Border Type: single, double, dash, dot, etc.
    + Alignment: left, right, center, decimal.
    + Column width. 
    + Text appearance: **bold**, *italics*, ~~strikethrough~~, underline and any combinations.
    + Font size. 
    + Text and border color (`r length(colors())` different colors named in `color()` function).
    + Special characters: any character in UTF-8 encoding (e.g. Greek, Symbol, Chinese, Japanese, Korean).
* Append several tables into one file. 
* Pagination.
* Built in raw data for validation.

