---
title: r2rtf Design Pattern
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
    toc_float: true
resource_files:
   - pdf/*.pdf
   - fig/*.png
   - html/*.html
   - html/*.png
   - docx/*.docx
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup, message=FALSE, echo=FALSE}
library(dplyr)
library(r2rtf)
```

## Overview 

> This document is for developers who would like to understand the internal of the `r2rtf` package.  

The r2rtf package is developed to standardize the approach to generate highly customized 
tables, listings and figures (TLFs) in RTF format and provide flexibility to 
customize table appearance for table title, subtitle, column header, footnote, and data source.

The r2rtf package is designed to enables pipes (`%>%`), 
the first argument are all `tbl` except `rtf_read_figure()` and `write_rtf()`.
A minimal example summarized the major steps in using r2rtf package to
create a table or listing after a data frame is ready.
  
```{r}
head(iris) %>% rtf_body() %>%                       # Step 1 Add attributes 
               rtf_encode() %>%                     # Step 2 Convert attributes to RTF encode 
               write_rtf(file = "rtf/ex-tbl.rtf")   # Step 3 Write to a .rtf file 
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/ex-tbl.pdf")
```

Similar step is used to create a figure. First we generate a figure in `png` format, then use `rtf_read_figure()` to read in the `png` file, and proceed to rtf generation.


```{r}
fig <- c("fig/fig1.png")
fig %>% rtf_read_figure() %>%                     # Step 1 Read in png file
         rtf_figure() %>%                      # Step 2 Add attributes 
         rtf_encode(doc_type = "figure") %>%   # Step 3 Convert attributes to RTF encode 
         write_rtf(file = "rtf/ex-fig.rtf")           # Step 4 Write to a .rtf file 
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/ex-fig.pdf")
```


## Data Structure and Attributes 

We explore `rtf_page()` to illustrate how attributes of a data frame is used. 
`rtf_page()` is a function to define the feature of a page in rtf document. 
For example, we need to set the page orientation, width, height, margin etc. 
The information is attached to an data frame using attributes. 

Therefore, the core part of `rtf_page()` is to assign attributes to the input data frame (`tbl`) as below.
Necessary input checking using `check_arg()`, `stopifnot()` and `match.arg()`, is provided in the function to provide informative error message with unexpected input.

```{r, eval = FALSE}
attr(tbl, "page")$orientation   <- orientation
attr(tbl, "page")$width         <- width
attr(tbl, "page")$height        <- height
attr(tbl, "page")$margin        <- margin
attr(tbl, "page")$border_color_first <- border_color_first
```

All attributes is saved in a logical way and summarized in the table below.  

```{r}
tbl <- head(iris) %>% rtf_page()
```

To access the attributes, `attr()` or `attributes()` function can be used. For example, to get page attributes 
in default setting, we can run code below. 

```{r}
data.frame(value = unlist(attr(tbl, "page")))
```


The table body attributes are more complicated than page attributes. Take `rtf_body()` as an example, it meets 
the user's need to take fully control of table appearance through parameters to customize table size, space, border 
type (e.g., single, double, dash, dot, etc.), color (e.g., 657 different colors named in `color()` function), 
line width, column width, row height, text format (e.g., bold, italics, strikethrough, underline and any combinations), 
font size, text color, alignment (e.g., left, right, center, decimal), etc. Format control can be at the cell, row, 
column, or table level. 

With the help of functions `obj_rtf_text()` and `obj_rtf_border()`, it is straightforward to pass the user inputs to 
text/border attributes. For example, parameter `text_justification = "c"` sets all text in the cells to be center adjusted 
and `text_justification = c("c", rep("l", 4))` sets the text in the first column to be center adjusted
and the texts in the remaining 4 columns to be left adjusted for a table with five columns.

In addition, `rtf_body()` initiates "page" attributes when color is used in any of table elements (border, text, etc.).

```{r}
head(iris) %>% rtf_body()
```

Similarly to `rtf_body()`, the package also includes 
`rtf_colheader()`, `rtf_footnote()`, `rtf_source()`, `rtf_subline()`, `rtf_text()`, `rtf_title()` and `rtf_figure()` to 
assign attributes to the certain component of the output. Below is a simple 
example of using the functions together for table/figure.

```{r, eval = FALSE}
head(iris) %>%                               # Step 1 Read in data frame
  rtf_title("iris") %>%                      # Step 2 Add title
  rtf_colheader("A | B | C | D | E") %>%     # Step 3 Add column header
  rtf_body() %>%                             # Step 4 Add attributes 
  rtf_footnote("This is a footnote") %>%     # Step 5 Add footnote 
  rtf_source("Source: xxx")                  # Step 6 Add data source
```

```{r, eval = FALSE}
"fig/tmp-fig.png" %>%
  rtf_read_figure() %>%                  # Step 1 Read PNG files from the file path
  rtf_title("title", "subtitle") %>%     # Step 2 Add title or subtitle
  rtf_footnote("footnote") %>%           # Step 3 Add footnote
  rtf_source("[datasource: mk0999]") %>% # Step 4 Add data source
  rtf_figure()                           # Step 5 Add figure attributes 
```

## RTF Encoding 

`rtf_encode()` wraps up internal functions in r2rtf package to translate all attributes attached to a data frame 
into [RTF syntax](http://www.snake.net/software/RTF/RTF-Spec-1.7.pdf). 
A good reference is [RTF pocket guide](https://www.oreilly.com/library/view/rtf-pocket-guide/9781449302047/).

All of the internal encoding functions in r2rtf package start with a prefix `as_rtf_`. 
For a table, `r2rtf:::rtf_encode_table()` is used internally to translate attribute to RTF syntax by using 
a set of `as_rtf_` functions. 

Likewise, `rtf_encode_list()` is used when we have multiple data frame with different data structures to be stacked, 
this is often seen in efficacy analysis. And `rtf_encode_figure()` is used for figure outputs.

`r2rtf:::as_rtf_colheader()` define the column header in a table or listing.
```{r, eval = FALSE}
head(iris) %>%
  rtf_body() %>%
  r2rtf:::as_rtf_colheader() %>%
  cat()
```

`r2rtf:::as_rtf_color()` define the color to be used in text or border.
```{r, eval = FALSE}
head(iris) %>%
  rtf_body(text_color = "red") %>%
  r2rtf:::as_rtf_color() %>%
  cat()
```

`r2rtf:::as_rtf_end()` define the end of rtf encode string.
```{r, eval = FALSE}
r2rtf:::as_rtf_end() %>% cat()
```

`r2rtf:::as_rtf_font()` define the font encode.
```{r, eval = FALSE}
r2rtf:::as_rtf_font() %>% cat()
```

`r2rtf:::as_rtf_footnote()` define the footnote encode.
```{r}
head(iris) %>%
  rtf_footnote("example") %>%
  r2rtf:::as_rtf_footnote() %>%
  cat()
```

`r2rtf:::as_rtf_init()` initiates rtf encode with 'English' as default language.
```{r, eval = FALSE}
r2rtf:::as_rtf_init() %>% cat()
```

`r2rtf:::as_rtf_margin()` define a page margin. 
```{r, eval = FALSE}
head(iris) %>%
  rtf_page() %>%
  r2rtf::as_rtf_margin() %>%
  cat()
```

`r2rtf:::as_rtf_new_page()` initiates new page encode. 
```{r, eval = FALSE}
r2rtf::as_rtf_new_page() %>% cat()
```

`r2rtf:::as_rtf_page()` define a page width (`\paperw`) and height(`\paperh`).
```{r, eval = FALSE}
head(iris) %>%
  rtf_page() %>%
  r2rtf:::as_rtf_page() %>%
  cat()
```

`r2rtf:::as_rtf_pageby()` define a page_by encoding when argument page_by is not NULL in `rtf_body()`.
```{r, eval = FALSE}
iris %>%
  rtf_body(page_by = "Species") %>%
  r2rtf:::as_rtf_pageby() %>%
  cat()
```

`r2rtf:::as_rtf_paragraph()` define a title/text attribute as paragraph.
```{r, eval = FALSE}
r2rtf:::as_rtf_paragraph(attr(head(iris) %>% rtf_title("example"), "rtf_title")) %>% cat()
```

`r2rtf:::as_rtf_source()` define the source encode.
```{r, eval = FALSE}
head(iris) %>%
  rtf_source("example") %>%
  r2rtf:::as_rtf_source() %>%
  cat()
```

`r2rtf:::as_rtf_subline()` define the subline encode.
```{r, eval = FALSE}
head(iris) %>%
  rtf_subline("example") %>%
  r2rtf:::as_rtf_subline() %>%
  cat()
```

`r2rtf:::as_rtf_table()` define the table encode.
```{r, eval = FALSE}
head(iris) %>%
  rtf_title("example") %>%
  r2rtf:::as_rtf_table() %>%
  cat()
```

`r2rtf:::as_rtf_title()` define the title encode.
```{r, eval = FALSE}
head(iris) %>%
  rtf_title("example") %>%
  r2rtf:::as_rtf_title() %>%
  cat()
```

In a minimal example, we can create an RTF file by combining different pieces of RTF code.
For simplicity, we only show the first two rows of `iris` data

```{r}
tbl <- iris[1:2, ] %>% rtf_body()

paste(
  r2rtf:::as_rtf_init(),
  r2rtf:::as_rtf_font(),
  r2rtf:::as_rtf_page(tbl),
  r2rtf:::as_rtf_margin(tbl),
  r2rtf:::as_rtf_table(tbl),
  "}",
  sep = "\n"
) %>% cat()
```

If we save the RTF code into an `.rtf` file. 
Microsoft Word or other RTF Viewer software can display the file properly.


## Save RTF File

After everything has been translated into RTF code using `rtf_encode`. It is a simple step to save all RTF code 
into an `.rtf` file. 

`write_rtf()` is a simple wrap up of `write` function, which exports a single RTF string into the output file.

A simple example is as in the overview section.

```{r, eval = FALSE}
head(iris) %>% rtf_body() %>% # Step 1 Add attributes
  rtf_encode() %>% # Step 2 Convert attributes to RTF encode
  write_rtf(file = "tmp.rtf") # Step 3 Write to a .rtf file
```

## Dictionary and Conversion

As mentioned earlier, functions in `dictionary.R` file contains the most commonly used font types, formats, border types, etc. 
All these attributes are mapped in a data frame which contains the element names and corresponding RTF encode. User is only 
allowed to key in whatever is defined in dictionary for the rtf_ functions, or error message will be provided as a result from `match_arg()`. 

```{r}
r2rtf:::font_type()
```

Superscripts, under scripts are taken care of by using `convert()` function to translate them into latex code and unicode. For example:

```{r}
r2rtf:::convert(c("^", "<="))
```

## Pageby 

It is commonly seen a variable displayed in headline as group category (e.g. baseline characteristic table). 
To achieve this, user first need to sort input data frame by page_by variable then define it in `rtf_body()`. 
Border and text attributes are controlled together with other variables in the vectors.

```{r, eval = FALSE}
iris %>%
  arrange(Species) %>%
  rtf_colheader("Sepal Length | Sepal Width | Petal Length | PetalWidth", ) %>%
  rtf_body(
    page_by = c("Species"),
    border_top = c(rep("", 4), c("single")),
    border_bottom = c(rep("", 4), c("single")),
    text_justification = c(rep("c", 4), c("l")),
    new_page = TRUE
  ) %>%
  rtf_encode() %>%
  write_rtf("pageby.rtf")
```


## Utility functions

* `check_args()`: Function for argument checking for types, length or dimension, this function is used for all export functions except `write_rtf()`.
* `match_arg()`: Function for argument verification on input values to see whether they match `dictionary()` defined values, 
this function is used for all export functions except `write_rtf`().
* 'footnote_source_space()`: Function to derive space adjustment, whose results could be used in `rtf_footnote()` and `rtf_source()` 
when indentation is defined by user.
    
 
