---
title: RTF Examples for Figures
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
devtools::load_all() # need to be updated after the package mature

library(dplyr)
```

The package allow user to embed multiple figures into one RTF document. The workflow can be summarized as

1. save figures into PNG format. (e.g. using `png()` or `ggplot2::ggsave()`).
1. read PNG files into R as binary file using `r2rtf::rtf_read_png()`.
1. add optional features using `r2rtf::title()`, `r2rtf::footnote()`, `r2rtf::source()`.
1. set up page and figure options using `r2rtf::rtf_figure()`.
1. encode rtf using `r2rtf::rtf_encode(type = "figure")`. (Note: it is important to set `type = "figure")` as the default is `type = "figure` to handle tables).
1. write rtf to a file using `r2rtf::write_rtf`.

A quick example to illustrate the workflow:
```{r}
# Define the path of figure
filename <- c("fig/fig1.png", "fig/fig2.png", "fig/fig3.png")

filename %>%
  rtf_read_png() %>%                     # read PNG files from the file path
  rtf_title("title", "subtitle") %>%     # add title or subtitle
  rtf_footnote("footnote") %>%           # add footnote
  rtf_source("[datasource: mk0999]") %>% # add data source
  rtf_figure() %>%                       # default seting of page and figure
  rtf_encode(doc_type = "figure") %>%        # encode rtf as figure
  write_rtf(file = "rtf/fig-simple.rtf") # write RTF to a file
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/fig-simple.pdf")
```

In `rtf_figure`, the features of page and figure can be set up: 

* Page orientation: `rtf_page`. 
* Figure height and width: `fig_heigth` and `fig_width`.

The figure height and width can be set up for each figure in a vector. 
The code below provides an example for these features.
```{r}
filename %>%
  rtf_read_png() %>%                                # read PNG files from the file path
  rtf_page(orientation = "landscape") %>%           # set page orientation
  rtf_title("title", "subtitle") %>%                # add title or subtitle
  rtf_footnote("footnote") %>%                      # add footnote
  rtf_source("[datasource: mk0999]") %>%            # add data source
  rtf_figure(
    fig_height = 3.5,                               # set figure height
    fig_width = c(6, 7, 8)                          # set figure width individually.
  ) %>% 
  rtf_encode(doc_type = "figure") %>%                   # encode rtf as figure
  write_rtf(file = "rtf/fig-landscape.rtf")         # write RTF to a file
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/fig-landscape.pdf")
```

