---
title: RTF Examples for Baseline Characteristics Tables
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(r2rtf)
library(dplyr)
library(tidyr)
```

# Example:

This example shows how to create a baseline characteristic table as below.

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/bs_example.pdf")
```

## Step 1: Define utility functions

We defined two utility functions for analysis purpose. 

- Summarize categorical variable 
```{r}
bs_count <- function(data, grp, var,
                     var_label = var,
                     decimal = 1,
                     total = TRUE) {
  data <- data %>% rename(grp = !!grp, var = !!var)
  coding <- levels(factor(data$grp))
  data <- data %>% mutate(grp = as.numeric(factor(grp)))

  res <- with(data, table(var, grp)) %>%
    as.data.frame() %>%
    mutate(grp = as.numeric(grp))

  if (total) {
    res_tot <- with(data, table(var)) %>%
      as.data.frame() %>%
      mutate(grp = 9999)
    res <- bind_rows(res, res_tot)
  }

  res <- res %>% rename(n = Freq)

  res <- res %>% mutate(pct = formatC(n / sum(n) * 100, digits = decimal, format = "f", flag = "0"))

  res$n <- as.character(res$n)

  res <- res %>%
    pivot_longer(cols = c(n, pct), names_to = "key", values_to = "value") %>%
    unite(keys, grp, key) %>%
    pivot_wider(names_from = keys, values_from = value) %>%
    mutate(var_label = var_label) %>%
    mutate(var = as.character(var))

  names(res) <- gsub("_n", "", names(res), fixed = TRUE)
  attr(res, "coding") <- coding

  res
}
```

- Summarize continuous variable 
```{r}
bs_continous <- function(data, grp, var,
                         var_label = var,
                         decimal = 1,
                         total = TRUE,
                         blank_row = FALSE) {
  data <- data %>% rename(grp = !!grp, var = !!var)
  coding <- levels(factor(data$grp))
  data <- data %>% mutate(grp = as.numeric(factor(grp)))

  res <- data %>%
    select(grp, var) %>%
    na.omit() %>%
    group_by(grp) %>%
    summarise(
      `Subjects with data` = n(),
      Mean = formatC(mean(var), digits = decimal, format = "f", flag = "0"),
      SD = formatC(sd(var), digits = decimal, format = "f", flag = "0"),
      Median = formatC(median(var), digits = decimal, format = "f", flag = "0"),
      Range = paste(range(var), collapse = " to ")
    )

  if (total) {
    res_tot <- data %>%
      select(grp, var) %>%
      na.omit() %>%
      summarise(
        `Subjects with data` = n(),
        Mean = formatC(mean(var), digits = decimal, format = "f", flag = "0"),
        SD = formatC(sd(var), digits = decimal, format = "f", flag = "0"),
        Median = formatC(median(var), digits = decimal, format = "f", flag = "0"),
        Range = paste(range(var), collapse = " to ")
      ) %>%
      mutate(grp = 9999)
    res <- bind_rows(res, res_tot)
  }

  res$"Subjects with data" <- as.character(res$"Subjects with data")

  res <- res %>%
    pivot_longer(cols = -grp, names_to = "key", values_to = "value") %>%
    mutate(key = factor(key, levels = c("Subjects with data", "Mean", "SD", "Median", "Range"))) %>%
    pivot_wider(names_from = grp, values_from = value) %>%
    mutate(var_label = var_label) %>%
    mutate(key = as.character(key)) %>%
    rename(var = key)

  if (blank_row) {
    res <- bind_rows(tibble(var_label = var_label), res)
  }

  res
}
```

## Step 2: Create data for rtf table

- Convert variable to factors to properly define display order

```{r}
# The code above define two utility function for baseline characteristic tables.

# Analysis Set
data(r2rtf_adsl)
ana <- r2rtf_adsl %>% subset(ITTFL == "Y")

ana <- ana %>% mutate(
  RACE = factor(
    RACE,
    c("WHITE", "BLACK OR AFRICAN AMERICAN", "AMERICAN INDIAN OR ALASKA NATIVE"),
    c("White", "Black", "Other")
  ),
  SEX = factor(
    SEX,
    c("F", "M"),
    c("Female", "Male")
  ),
  AGEGR1 = factor(
    AGEGR1,
    levels = c("<65", "65-80", ">80")
  )
)

# Build Data for r2rtf
bs_tb <- bind_rows(
  bs_count(ana, "TRT01AN", "SEX", "Gender"),
  bs_count(ana, "TRT01AN", "AGEGR1", "Age (Years)"),
  bs_continous(ana, "TRT01AN", "AGE", "Age (Years)", blank_row = TRUE),
  bs_count(ana, "TRT01AN", "RACE", "Race")
) %>% mutate(
  var_label = factor(var_label, levels = c("Gender", "Age (Years)", "Race"))
)

bs_tb[is.na(bs_tb)] <- "" # convert NA to blank string.
```

```{r}
knitr::kable(bs_tb)
```

## Step 3: Define table format

- page_by feature was used to display 3 parts (Gender, Age, Race) clearly
  + if using `new_page = TRUE` in `rtb_body()`, the 3 parts will be outputted into 3 pages
  + `page_by` variable should not be in `rtf_colheader()`. However, it should be in `rtf_body()`

```{r}
bs_rtf <- bs_tb %>%
  rtf_page(width = 9.5) %>%
  rtf_title("Demographic and Anthropometric Characteristics", "ITT Subjects") %>%
  rtf_colheader(" | Placebo | Drug Low Dose | Drug High Dose | Total ",
    col_rel_width = c(3, rep(2, 4))
  ) %>%
  rtf_colheader(" | n | (%) | n | (%) | n | (%) | n | (%) ",
    col_rel_width = c(3, rep(c(1.2, 0.8), 4)),
    border_top = c("", rep("single", 8)),
    border_left = c("single", rep(c("single", ""), 4))
  ) %>%
  rtf_body(
    page_by = "var_label",
    col_rel_width = c(3, rep(c(1.2, 0.8), 4), 3),
    text_justification = c("l", rep("c", 8), "l"),
    text_format = c(rep("", 9), "b"),
    border_left = c("single", rep(c("single", ""), 4), "single"),
    border_top = c(rep("", 9), "single"),
    border_bottom = c(rep("", 9), "single"),
  ) %>%
  rtf_footnote("This is a footnote") %>%
  rtf_source("Source: xxx")
```

## Step 4: Output

```{r}
# Output .rtf file
bs_rtf %>%
  rtf_encode() %>%
  write_rtf("rtf/bs_example.rtf")
```
