---
title: RTF Examples for Figures
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(r2rtf)
library(dplyr)
library(tidyr)
```

## Examples

This example shows how to create and embed figures into an RTF file as below.

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/fig-simple.pdf")
```

Below is an example with adjusted page orientation, figure height and width.

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/fig-landscape.pdf")
```

### Overall workflow

The package allow user to embed multiple figures into one RTF document. 
The supported format is listed as below. 

```{r}
r2rtf:::fig_format()
```

By using `png` file as an example, the workflow can be summarized as:

1. Save figures into PNG format. (e.g. using `png()` or `ggplot2::ggsave()`).
1. Read PNG files into R as binary file using `r2rtf::rtf_read_figure()`.
1. Add optional features using `r2rtf::rtf_title()`, `r2rtf::rtf_footnote()`, `r2rtf::rtf_source()`.
1. Set up page and figure options using `r2rtf::rtf_page()`, `r2rtf::rtf_figure()`.
1. Encode rtf using `r2rtf::rtf_encode(doc_type = "figure")`. (Note: it is important to set `doc_type = "figure"` as the default is `doc_type = "table"` to handle tables).
1. Write rtf to a file using `r2rtf::write_rtf()`.

> For `emf` format, one may use the R package `devEMF` to create the figure. 

#### Simple Example

```{r}
# Define the path of figure
filename <- c("fig/fig1.png", "fig/fig2.png", "fig/fig3.png")

filename %>%
  rtf_read_figure() %>%                  # read PNG files from the file path
  rtf_title("title", "subtitle") %>%     # add title or subtitle
  rtf_footnote("footnote") %>%           # add footnote
  rtf_source("[datasource: mk0999]") %>% # add data source
  rtf_figure() %>%                       # default setting of page and figure
  rtf_encode(doc_type = "figure") %>%    # encode rtf as figure
  write_rtf(file = "rtf/fig-simple.rtf") # write RTF to a file
```

#### Example with features

Features of page and figure can be set up in `rtf_page` and `rtf_figure` respectively: 

* Page orientation: `orientation` argument in `rtf_page`. 
* Figure height and width: `fig_heigth` and `fig_width` arguments in `rtf_figure`.

The figure height and width can be set up for each figure in a vector. 
The code below provides an example for these features.

```{r}
filename %>%
  rtf_read_figure() %>%                             # read PNG files from the file path
  rtf_page(orientation = "landscape") %>%           # set page orientation
  rtf_title("title", "subtitle") %>%                # add title or subtitle
  rtf_footnote("footnote") %>%                      # add footnote
  rtf_source("[datasource: mk0999]") %>%            # add data source
  rtf_figure(
    fig_height = 3.5,                               # set figure height
    fig_width = c(6, 7, 8)                          # set figure width individually.
  ) %>% 
  rtf_encode(doc_type = "figure") %>%               # encode rtf as figure
  write_rtf(file = "rtf/fig-landscape.rtf")         # write RTF to a file
```
