---
title: "RTF Examples for Figures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example-figure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(dplyr)
library(r2rtf)
```

The package allow user to embed multiple figures into one RTF document. The workflow can be summarized as

1. save figures into PNG format. (e.g. using `png()` or `ggplot2::ggsave()`).
1. read PNG files into R as binary file using `r2rtf::rtf_read_png()`.
1. add optional features using `r2rtf::title()`, `r2rtf::footnote()`, `r2rtf::source()`.
1. set up page and figure options using `r2rtf::rtf_figure()`.
1. encode rtf using `r2rtf::rtf_encode(type = "figure")`. (Note: it is important to set `type = "figure")` as the default is `type = "figure` to handle tables).
1. write rtf to a file using `r2rtf::write_rtf`.

A quick example to illustrate the workflow:
```{r, eval = FALSE}
# Define the path of figure
filename <- c("fig/fig1.png", "fig/fig2.png", "fig/fig3.png")

filename %>%
  rtf_read_png() %>% # read PNG files from the file path
  rtf_title("title", "subtitle") %>% # add title or subtitle
  rtf_footnote("footnote") %>% # add footnote
  rtf_source("[datasource: mk0999]") %>% # add data source
  rtf_figure() %>% # default seting of page and figure
  rtf_encode(type = "figure") %>% # encode rtf as figure
  write_rtf(file = "rtf/fig-simple.rtf") # write RTF to a file
```

In `rtf_figure`, the features of page and figure can be set up: 

* Page direction: `rtf_figure`. 
* Page margin: `doctype`.
* Page height and width: `page_height` and `page_width`.
* Figure height and width: `fig_heigth` and `fig_width`.

The figure height and width can be set up for each figure in a vector. The code below provides an example for these features.
```{r, eval = FALSE}
filename %>%
  rtf_read_png() %>% # read PNG files from the file path
  rtf_title("title", "subtitle") %>% # add title or subtitle
  rtf_footnote("footnote") %>% # add footnote
  rtf_source("[datasource: mk0999]") %>% # add data source
  rtf_figure(
    page_height = 8.5, # page height
    page_width = 11, # page width
    orientation = "landscape", # page direction
    doctype = "narrow", # set page margin as narrow i.e. 0.5 inch at each side
    fig_height = 5, # set figure height
    fig_width = c(6, 7, 8)
  ) %>% # set figure width individually.
  rtf_encode(type = "figure") %>% # encode rtf as figure
  write_rtf(file = "rtf/fig-landscape.rtf") # write RTF to a file
```

