---
title: RTF Examples for Tables with Sublineby, Pageby, and Groupby Features
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(r2rtf)
library(dplyr)
library(tidyr)
```

# Overview

In `rtf_body`, `r2rtf` provided three advanced arguments to customize table layout. 

- `page_by`: The variable is used as section header.   
- `group_by`: The variable only display when it is first appeared. 
- `subline_by`: The variable is used as page subline header.

If one or more of the arugments are used, they need to be properly ordered 
in `subline_by`, `page_by` and `group_by`. 

For `page_by`, the argument `new_page` and `pageby_header` and `pageby_row` further 
customize the table layout. Details can be found in the `rtf_body` documentation. 

We provided three examples below to illustrate the idea. 

# Example 1: Pageby 

We used `page_by` variable to illustrate how to create a simple disposition table. 

> Note: This example also illustrate how to avoid row header by setting corresponding value 
> in `page_by` variable to `"-----"`.

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/pageby-disposition.pdf")
```

## Step 1: Create data for RTF table

```{r}
data(r2rtf_adsl)
adsl <- r2rtf_adsl

# randomized
row1 <- adsl %>%
  subset(ITTFL == "Y") %>%
  group_by(TRT01P) %>%
  count()

# treated
row2 <- adsl %>%
  subset(SAFFL == "Y") %>%
  group_by(TRT01P) %>%
  summarise(trtn = n())

# completed study
row3 <- adsl %>%
  subset(SAFFL == "Y" & DCREASCD == "Completed") %>%
  group_by(TRT01P) %>%
  summarise(cmpln = n()) 

# discontinued study
row4 <- adsl %>%
  subset(SAFFL == "Y" & DCREASCD != "Completed") %>%
  group_by(TRT01P) %>%
  summarise(dcn = n()) 

# discontinuation reason
reas <- adsl %>%
  subset(SAFFL == "Y" & DCREASCD != "Completed") %>%
  group_by(TRT01P, DCREASCD) %>%
  summarise(reasn = n()) 

ds <- rbind(
  row1 %>% merge(row2, by = "TRT01P") %>%
    merge(row3, by = "TRT01P") %>%
    merge(row4, by = "TRT01P") %>%
    pivot_longer(2:5, names_to = "DCREASCD", values_to = "n") %>%
    pivot_wider(names_from = TRT01P, values_from = "n"),

  reas %>% pivot_wider(names_from = TRT01P, values_from = c("reasn"))
) %>%
  mutate(
    DCREASCD = ifelse(DCREASCD == "n", "Participants randomized",
      ifelse(DCREASCD == "trtn", "Participants treated",
        ifelse(DCREASCD == "cmpln", "Participants completed",
          ifelse(DCREASCD == "dcn", "Participants discontinued", paste0("   ", DCREASCD))
        )
      )
    ),
    pagebyvar = ifelse(substr(DCREASCD, 1, 5) == "Parti", "-----", "Discontinued reason")
  ) %>%

  # pagebyvar will be assigned to page_by argument in rtf_body function later.
  # then table will be grouped by single cell row with pagebyvar's value in it.
  # if pagebyvar's value is "-----", the single cell row will be removed from table.

  select(pagebyvar, DCREASCD, "Xanomeline High Dose", "Xanomeline Low Dose", Placebo)

ds[is.na(ds)] <- 0
```

```{r}
knitr::kable(head(ds))
```

## Step 2: Define table format

```{r}
ds_tbl <- ds %>%
  
  rtf_title("Disposition of Participants", "(ITT Population)") %>%
  
  rtf_colheader(" | Xanomeline High Dose |Xanomeline Low Dose | Placebo",
    col_rel_width = c(3, 2, 2, 2)
  ) %>%
  
  rtf_colheader(" | n | n | n ",
    border_top = c("", rep("single", 3)),
    col_rel_width = c(3, rep(2, 3))
  ) %>%
  
  # the table will be grouped by single cell row with pagebyvar's value in it.
  # if pagebyvar's value is "-----", the single cell row will be removed from table.
  rtf_body(
    page_by = "pagebyvar",
    col_rel_width = c(1, 3, 2, 2, 2),
    text_justification = c("l", "l", rep("c", 3)),
    text_format = c("b", rep("", 4)),
    border_top = c("single", rep("", 4)),
    border_bottom = c("single", rep("", 4)),
  ) %>%
  
  rtf_footnote(c("This is footnote")) %>%

  rtf_source("Source:  [Study MK9999P001: adam-adsl]")
```

## Step 3: Output

```{r}
# Output .rtf file
ds_tbl %>% 
  rtf_encode() %>% 
  write_rtf("rtf/pageby-disposition.rtf")
```

# Example 2: Pageby with `pageby_row = "first_row"` 

We used `page_by` variable and `pageby_row = "first_row"` to illustrate 
how to create a simple adverse events table. 

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/pageby-firstrow-ae.pdf")
```

## Step 1: Create data for RTF table 

```{r, message=FALSE}
# Read in and merge r2rtf_adsl and r2rtf_adae data
data(r2rtf_adsl)
data(r2rtf_adae)

ana <- r2rtf_adsl %>% 
       subset(SAFFL == "Y") %>%     
       select(USUBJID, TRT01AN, TRT01A) 

# inner join adsl to bring in TRT01AN
aedata <- inner_join(ana, r2rtf_adae, by="USUBJID") %>% 
          filter( AEDECOD != '' & SAFFL =='Y' & AESEV == 'SEVERE') %>% 
          mutate( AEDECOD = tolower(AEDECOD),
                  AEBODSYS = tolower(AEBODSYS)                  
           )

# participants part
overall <- ana %>% 
  group_by(TRT01AN) %>% 
  summarise(n = n_distinct(USUBJID)) %>% 
  mutate(AEDECOD = 'Participants in population')

withAE <- aedata %>% 
  group_by(TRT01AN) %>% 
  summarise(withAE = n_distinct(USUBJID))%>% 
  mutate(AEDECOD = '  with one or more severe AE') %>% 
  rename(n = withAE)

woAE <- left_join(overall, withAE, by = 'TRT01AN', suffix = c(".x", ".y")) %>% 
  mutate(woAE = n.x-n.y,
         AEDECOD = '  with no severe AE') %>% 
  select(TRT01AN, AEDECOD, woAE) %>% 
  rename(n = woAE)

part1 <- bind_rows(overall, withAE, woAE) %>% 
  pivot_wider(names_from = TRT01AN, values_from = n) %>% 
  mutate(AEBODSYS = '  a')
 
# AE part
bodsys <- aedata %>% 
  group_by(TRT01AN, AEBODSYS) %>% 
  summarise(n = n_distinct(USUBJID)) 

decod <- aedata %>% 
  group_by(TRT01AN, AEBODSYS, AEDECOD) %>% 
  summarise(n = n_distinct(USUBJID)) %>% 
  arrange(AEBODSYS, AEDECOD)

part2 <- bind_rows(bodsys, decod) %>% 
  pivot_wider(names_from = TRT01AN, values_from = n, values_fill = 0) %>% 
  arrange(AEBODSYS, !is.na(AEDECOD)) %>% 
  mutate(AEDECOD = ifelse(is.na(AEDECOD), AEBODSYS, paste('  ', AEDECOD)))

blank <- tibble(AEBODSYS = '  a')

apr0ae <- bind_rows(part1, blank, part2)  
 
```

```{r}
knitr::kable(head(apr0ae))
```

## Step 2: Define table format

```{r}
apr0ae_rtf <-  apr0ae %>%
  rtf_page(orientation = "landscape") %>% 
  rtf_title(c("Participants with Severe Adverse Events","(Incidence \\geq 0% in One or More Treatment Groups)"), "(ASaT Population)") %>%

  rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose  ", 
                col_rel_width = c(5, 2, 2, 2),
                border_bottom = c("", rep("single", 3))) %>%

  rtf_colheader(" | n |  n  | n  ",
                border_top    = c("", rep("single", 3)),
                border_bottom = "single",
                col_rel_width = c(5, rep(2, 3)),  
                ) %>%

  rtf_body(col_rel_width = c(5, rep(2, 3), 1),
           text_justification = c("l", "c", "c", "c", "l"),
           border_first = "",  
           page_by = "AEBODSYS",
           pageby_row = "first_row", 
           ) %>% 

  rtf_footnote(c("This is footnote")) %>% 

  rtf_source("Source:  [Study MK9999P001: adam-adae]")
```

## Step 3: Output

```{r}
# Output .rtf file
apr0ae_rtf %>% 
  rtf_encode() %>% 
  write_rtf("rtf/pageby-firstrow-ae.rtf")
```

# Example 3: sublineby, pageby, and groupby features

We used `page_by`, `subline_by` and `group_by` to illustrate 
how to create a simple adverse events listing. 

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/pageby-ae-listing.pdf")
```

## Step 1: Create data for RTF table

```{r}
data(r2rtf_adae)

ae_t1 <- r2rtf_adae[200:260, ] %>%
  mutate(
    SUBLINEBY = paste0(
      "Trial Number: ", STUDYID,
      ", Site Number: ", SITEID
    ),
    SUBJLINE = paste0(
      "Subject ID = ", USUBJID,
      ", Gender = ", SEX,
      ", Race = ", RACE,
      ", AGE = ", AGE, " Years",
      ", TRT = ", TRTA
    ),
    # create a subject line with participant's demographic information.
    # this is for page_by argument in rtf_body function
    AEDECD1 = tools::toTitleCase(AEDECOD), # propcase the AEDECOD
    DUR = paste(ADURN, ADURU, sep = " ")
  ) %>% # AE duration with unit
  select(USUBJID, ASTDY, AEDECD1, DUR, AESEV, AESER, AEREL, AEACN, AEOUT, TRTA, SUBJLINE, SUBLINEBY) # display variable using this order
```

```{r}
knitr::kable(head(ae_t1, 2))
```

## Step 2: Define table format

```{r}
ae_tbl <- ae_t1 %>%
  # It is important to order variable properly. 
  arrange(SUBLINEBY, TRTA, SUBJLINE, USUBJID, ASTDY) %>%
  
  rtf_page(orientation = "landscape", col_width = 9) %>%
  
  rtf_page_header() %>%
  
  rtf_page_footer(text = "CONFIDENTIAL") %>%
  
  rtf_title("Listing of Subjects With Serious Adverse Events",
            "ASaT") %>%
  
  rtf_colheader("Subject| Rel Day | Adverse | | | | |Action| |",
    col_rel_width = c(2.5, 2, 4, 2, 3, 2, 3, 2, 5)
  ) %>%
  
  rtf_colheader("ID| of Onset | Event | Duration | Intensity | Serious |
                Related | Taken| Outcome",
    border_top = "",
    col_rel_width = c(2.5, 2, 4, 2, 3, 2, 3, 2, 5)) %>%
  
  rtf_body(
    col_rel_width = c(2.5, 2, 4, 2, 3, 2, 3, 2, 5, 1, 1, 1),
    text_justification = c("l", rep("c", 8), "l", "l", "l"),
    text_format = c(rep("", 9), "b", "", "b"),
    border_top = c(rep("", 9), "single", "single", "single"),
    border_bottom = c(rep("", 9), "single", "single", "single"),
    subline_by = 'SUBLINEBY',
    page_by = c("TRTA", "SUBJLINE"),
    group_by = c("USUBJID", "ASTDY")
  ) %>%
  
  rtf_footnote(c("This is footnote 1")) %>% 

  rtf_source("Source:  [Study MK9999P001: adam-adae]")
```

## Step 3: Output

```{r}
# Output .rtf file
ae_tbl %>% 
  rtf_encode() %>% 
  write_rtf("rtf/pageby-ae-listing.rtf")
```


