---
title: RTF Examples for Controlling Text and Paragraph Details
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(r2rtf)
library(dplyr)
```

This vignette documents how to use `rtf_text` and `rtf_paragraph` to customize text and paragraph in details. All output tables are saved in the `vignettes/rtf` folder and assembled in `vignettes/r2rtf_examples.docx` document. 

```{r}
text <- paste(rep("Sample Text", 20), collapse = " ")
```

```{r}
text <- rep(text, 5)
```

## Paragraph 

### Paragraph Alignment 

Paragraph alignment supports four types. 

```{r}
r2rtf:::justification() %>% subset(type %in% c("l", "c", "r", "j"))
```

This example display text in different alignment methods.

```{r}
res <- r2rtf:::rtf_paragraph(r2rtf:::rtf_text(text),
  justification = c("l", "c", "r", "j")
)
r2rtf:::write_rtf_para(res, "rtf/para-justification.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/para-justification.pdf")
```

### Indent 

First line, left, and right indent can be controlled. One can also use a negative number to have the text "outdent".

```{r}
res <- r2rtf:::rtf_paragraph(r2rtf:::rtf_text(text),
  indent_first = c(1000, 0, -1000),
  indent_left = c(500, -500),
  indent_right = 500
)
r2rtf:::write_rtf_para(res, "rtf/para-indent.rtf")
```


```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/para-indent.pdf")
```

### Line Space

Different types of line space can be controlled.

```{r}
r2rtf:::spacing()[, 1:2]
```

This example show different line spaces in paragraph. 

```{r}
res <- r2rtf:::rtf_paragraph(r2rtf:::rtf_text(text),
  space = c(1, 2, 1.5)
)
r2rtf:::write_rtf_para(res, "rtf/para-line-space.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/para-line-space.pdf")
```

### Paragraph Space

```{r}
res <- r2rtf:::rtf_paragraph(r2rtf:::rtf_text(text),
  space_before = c(50, 180),
  space_after = c(180, 50)
)
r2rtf:::write_rtf_para(res, "rtf/para-space.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/para-space.pdf")
```

### New Page

This example add page break before paragraphs. 
```{r}
res <- r2rtf:::rtf_paragraph(r2rtf:::rtf_text(text)[1:2],
  new_page = TRUE
)
r2rtf:::write_rtf_para(res, "rtf/para-page.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/para-page.pdf")
```

## Text

### Font Size

```{r}
res <- r2rtf:::rtf_paragraph(r2rtf:::rtf_text(text,
  font_size = 8:12
))
r2rtf:::write_rtf_para(res, "rtf/text-font-size-1.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/text-font-size-1.pdf")
```

### Text Format

```{r}
res <- r2rtf:::rtf_paragraph(r2rtf:::rtf_text(text,
  format = c("b", "i", "bi", "^", "_", "u", "s")
))
r2rtf:::write_rtf_para(res, "rtf/text-format-1.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/text-format-1.pdf")
```

### Text Font Type

```{r}
res <- r2rtf:::rtf_paragraph(r2rtf:::rtf_text(text,
  font = 1:10
))
r2rtf:::write_rtf_para(res, "rtf/text-font-type.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/text-font-type.pdf")
```

### Text Color

```{r}
res <- r2rtf:::rtf_paragraph(r2rtf:::rtf_text(text,
  color = c("red", "gold", "black", "orange", "blue")
))
r2rtf:::write_rtf_para(res, "rtf/text-color.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/text-color.pdf")
```

### Text Background Color

```{r}
res <- r2rtf:::rtf_paragraph(r2rtf:::rtf_text(text,
  color = "white",
  background_color = c("red", "gold", "black", "orange", "blue")
))
r2rtf:::write_rtf_para(res, "rtf/text-background-color.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/text-background-color.pdf")
```

### Combine Text in Different Format

This example call `rtf_text` multiple times to combine a text.

```{r}
res <- r2rtf:::rtf_paragraph(paste0(
  r2rtf:::rtf_text("3.5"),
  r2rtf:::rtf_text("\\dagger", format = "^"),
  r2rtf:::rtf_text("\\line red ", color = "red"),
  r2rtf:::rtf_text("highlight", background_color = "yellow")
))
r2rtf:::write_rtf_para(res, "rtf/text-combine1.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/text-combine1.pdf")
```

This example call `rtf_text` one time to combine a text.

```{r}
text <- c(3.5, "\\dagger", "\\line red ", "highlight")
format <- c("", "^", "", "")
color <- c("black", "black", "red", "black")
background_color <- c("white", "white", "white", "yellow")

res <- r2rtf:::rtf_paragraph(
  paste(
    r2rtf:::rtf_text(text,
      format = format,
      color = color,
      background_color = background_color
    ),
    collapse = ""
  )
)
r2rtf:::write_rtf_para(res, "rtf/text-combine2.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/text-combine2.pdf")
```

### Inline Formatting

This example provide an inline formatting options with superscript and subscript. It is important to note the location of `{}` is before the special character.

```{r}
text <- c("X{_1} = \\alpha{^2} + \\beta{^\\dagger}")
res <- r2rtf:::rtf_paragraph(r2rtf:::rtf_text(text))
r2rtf:::write_rtf_para(res, "rtf/text-combine3.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/text-combine3.pdf")
```
