---
title: "File Format Conversion"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
library(r2rtf)
```

# Overview

`r2rtf` had an experimental internal function `r2rtf:::rtf_convert_format` to 
convert RTF format to PDF, DOCX or HTML. 
This feature requires [libreoffice7.1](https://www.libreoffice.org/download/download/) and only 
be tested in a Linux environment. 

By default, `r2rtf:::rtf_convert_format` convert RTF file to PDF format.
The PDF file will be saved in the same folder with same file name by default.

In example below, `ae_example.pdf` will be created in the same folder. 

```{r, eval = FALSE}
r2rtf:::rtf_convert_format(input = "rtf/ae_example.rtf")
```

- Multiple files can be provided as input.

```{r, eval = FALSE}
r2rtf:::rtf_convert_format(input = list.files("rtf", pattern = "*.rtf"))
```

- The `format` argument allow use to specify output format as `pdf`, `docx`, or `html`. 

```{r, eval = FALSE}
r2rtf:::rtf_convert_format(input = "rtf/ae_example.rtf", format = "html")
```

## Limitations 

There are 2 known limitations after RTF file convert to HTML files:   

- table border issue.
  - "dash", "small dash", "dot dash", and "dot dot" border type are not supported 

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/border-type.pdf")
```

````{=html}
```{r, echo=FALSE, results='asis'}
xfun::file_string('html/border-type.html')
```
````

- left and right adjusted paragraph outside of table. 
  - e.g. title, subtitle, subline, footnote, or source. 

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("pdf/intro-ae16.pdf")
```

````{=html}
```{r, echo=FALSE, results='asis'}
xfun::file_string("html/intro-ae16.html")
```
````
